---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# heartbeatr

<!-- badges: start -->
<!-- badges: end -->

A simple workflow to process data collected with PULSE systems (www.electricblue.eu/pulse).

## Installation

You can install the development version of **heartbeatr** from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("coastalwarming/heartbeatr")
```

## Example

List PULSE files to be read:

```{r paths}
library(heartbeatr)
# but make sure they correspond to a single experiment/device
# ...here we use the package's example data
paths <- pulse_example("RAW_original_")
paths
```

There are two ways to read and process those data:

```{r process}
# step by step
pulse_data_sub   <- pulse_read(paths, msg = FALSE)
pulse_data_split <- pulse_split(
   pulse_data_sub,
   window_width_secs = 30,
   window_shift_secs = 60,
   min_data_points = 0.8, 
   msg = FALSE
   )
pulse_data_split <- pulse_optimize(pulse_data_split, target_freq = 40, bandwidth = 0.2)
heart_rates <- pulse_heart(pulse_data_split, msg = FALSE)

# or calling a single wrapper function
heart_rates <- PULSE(
  paths,
  discard_channels  = paste0("s", 5:10), # channels s5 to s10 are empty in the example data
  window_width_secs = 30,
  window_shift_secs = 60,
  min_data_points   = 0.8,
  target_freq = 40,
  bandwidth   = 0.2,
  msg = FALSE
  )
```

Once processed, PULSE data is stored as a tibble with a heart rate frequency value for each channel/split window.

```{r check}
heart_rates
```

You can easily use parallel computing with **heartbeatr** - just configure your R session properly **BEFORE** applying the PULSE workflow:

```{r parallel}
# this shows how your session is currently configured 
#   (typically defaults to "sequential", i.e., not parallelized)
future::plan()

# to make use of parallel computing (highly recommended)
future::plan("multisession")
future::plan()
```

The raw data underlying the heart rate frequency estimate (hz) can be inspected:

```{r plot1}
pulse_plot_raw(heart_rates, ID = "limpet_1", i = 5, range = 2) 
# the 5th split window for channel "limpet_1" (the target --> i = 5) is shown in the center
# - 2 more windows are shown before and after the target
# - red dots show where the algorithm detected a peak
```

A quick overview of the result of the analysis:

```{r plot2}
pulse_plot_all(heart_rates)
```

The number of data points can be reduced:

```{r plot3}
heart_rates_binned <- pulse_summarise(heart_rates, fun = mean, span_mins = 3, min_data_points = 0.8)
pulse_plot_all(heart_rates_binned)
```
That's it. Have fun!
